Mastery Learning in Practice:\\ A Descriptive Analysis of Data from the Cognitive Tutor Algebra I Effectiveness Trial

### 2017-10-10 paper.Rnw
more stuff

Here we extend the previous findings to investigate patterns of change in the sequence of topics presented to students. As noted earlier, schools could use a standard curriculum, a customized curriculum, or switch from one curriculum to another. Each curriculum consisted of a sequence of units, which was comprised of a sequence of sections. For example, the standard CTAI Algebra I curriculum in Year 1 comprised of 40 units covering approximately 170 sections. The
number of sections within each unit varied. By default, the software operates by automatically moving students from section to section based on the sequence of topics defined by the curriculum they were currently enrolled in. In this software-controlled sequencing, students ideally spend the time necessary to learn the material of a section, are judged by the software to have mastered the material, and then “graduate” to the next section. Students who exhaust a section's material without mastering its skills are “promoted” to the next section. Students left uninterrupted follow the sequence of topics defined by their curriculum, ideally graduating from topics by showing mastery. However, teachers could modify a student’s path within the curriculum. They could reassign students from their current sections to other sections earlier or later in the intended sequence, including sections they had worked on previously.
This study attempts to elucidate teachers' goals in reassigning students. One hypothesis is the need for teachers push ahead students who were falling behind, i.e. to reassign them from sections on which they were struggling, to allow them catch up with the rest of the class. Another hypothesis is that teachers sought to push students past easier sections to begin working on more relevant topics for them. A third hypothesis is that teachers needed to cover certain topics in preparation for an upcoming state exams, and might have reassigned groups of students all at the same time to cover topics that might otherwise not have been covered. This hypothesis may lead to an increase in reassignments as the exam approaches. Overall, understanding patterns of reassignment is the focus of this paper, along with insights into how teachers were using the software component of CTAI.

@@

omar <- par()$oma

par(oma=c(0,2,0,0),mar=c( 3,.5,2,0))
par(mfrow=c(1,9))
for(st in levels(data$State)){
#    cwy <- currWorkedSt%>%filter(State==st)%>%group_by(Curriculum,overall)%>%summarize(`Yr 1`=p[Year=='Year 1'],`Yr 2`=p[Year=='Year 2'])
    cwy <- data%>%filter(!is.na(Curriculum) & State==st)%>%group_by(Curriculum,overall)%>%
    summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
    for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
    cwy <- cwy[nrow(cwy):1,]
    barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',75,NA),main=st,yaxt='n')
    if(st==levels(data$State)[1]) axis(2,at=seq(0,1,0.1),labels=paste0(seq(0,100,10),'\\%'))
    #box()
}
cwy <- data%>%filter(!is.na(Curriculum))%>%group_by(Curriculum,overall)%>%
    summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
cwy <- cwy[nrow(cwy):1,]
barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',75,NA),main='Overall',yaxt='n')

plot(1:10,type='n',xaxt='n',yaxt='n',bty='n')
legend(-1.5,7.5, c("Bridge-to-\n Algebra",'Algebra I','$>$Alg. I','','Standard','Cust.'),
       fill=c(COLS,NA,'#010101','#010101'),border=rep('white',6),
       angle=c(NA,NA,NA,NA,NA,45),density=c(NA,NA,NA,NA,NA,75),xpd = TRUE, bty = "n",inset=c(0,0))

par(oma=omar)
@
\caption{Percentage of worked problems coming from various levels
  (denoted by color, with Algebra II and Geometry bundled as ``$>$Algebra I''), from
  standard and customized variants, denoted by shading.}
\label{fig:curricula}
\end{figure}

@@

\begin{figure}
<<usageTime, fig.height=3,fig.width=6,dependson=c('data','setup','usageMedians')>>=

print(timeStateYear <- ggplot(secByStud2,aes(Yr,time))+geom_boxplot(outlier.shape=NA)+
          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.time), width=0.2)+
          facet_grid(~state)+coord_cartesian(ylim=c(0,110))+labs(x='',y='Hours on CT Software per Student'))
                                        #ggsave('timeStatYear.jpg',width=6,height=3)

@@

<<supplementalFigs1,include=FALSE,dependson=c('data','setup','usageMedians')>>=
secStateYear <- ggplot(secByStud2,aes(Yr,numSec))+geom_boxplot(outlier.shape=NA)+
          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.sec), width=0.2)+
          facet_grid(~state)+labs(x='',y='# of Sections Worked per Student')+coord_cartesian(ylim=c(0,200))
ggsave('secStateYear.jpg',width=6,height=3)

probStateYear <- ggplot(secByStud2,aes(Yr,nprob))+geom_boxplot(outlier.shape=NA)+
          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.prob), width=0.2)+
          facet_grid(~state)+labs(x='',y='# of Problems Worked per Student')+coord_cartesian(ylim=c(0,2000))
ggsave('probStateYear.jpg',width=6,height=3)

@@

\begin{figure}
<<unitsWorked,dependson=c('usageMedians','data','setup'),fig.width=6,fig.height=3>>=
secByStud2 <-
  secByStud2 %>%
  group_by(state,Yr) %>%
  mutate(outlier = numUnit > quantile(numUnit,0.75,na.rm=TRUE) + IQR(numUnit,na.rm=TRUE) * 1.5) %>%
  ungroup

print(unitsStateYear <- ggplot(secByStud2,aes(Yr,numUnit))+geom_boxplot(outlier.shape=NA)+geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier), width=0.2)+facet_grid(~state)+labs(x='',y='# Units Worked Per Student')+coord_cartesian(ylim=c(0,55)))
@
\caption{Boxplots of the number of units of Cognitive Tutor
  software each student worked, by year and state. Students
  working more than 55 units (\Sexpr{sum(secByStud[['numUnit']]>55,na.rm=TRUE)}) of \Sexpr{nrow(secByStud)}) and students
  with no usage data (\Sexpr{sum(is.na(secByStud[['numUnit']]))})
  were excluded.}
\label{fig:unitsByStud}
\end{figure}

@@

\section{Mastery Learning: Working Units in Order---Or Not}

@@

>>>>

Here we extend the previous findings to investigate patterns of change in the sequence of topics presented to students. As noted earlier, schools could use a standard curriculum, a customized curriculum, or switch from one curriculum to another. Each curriculum consisted of a sequence of units, which was comprised of a sequence of sections. For example, the standard CTAI Algebra I curriculum in Year 1 comprised of 40 units covering approximately 170 sections. The
number of sections within each unit varied.
This study attempts to elucidate teachers' goals in reassigning students. One hypothesis is the need for teachers push ahead students who were falling behind, i.e. to reassign them from sections on which they were struggling, to allow them catch up with the rest of the class. Another hypothesis is that teachers sought to push students past easier sections to begin working on more relevant topics for them. A third hypothesis is that teachers needed to cover certain topics in preparation for an upcoming state exams, and might have reassigned groups of students all at the same time to cover topics that might otherwise not have been covered. This hypothesis may lead to an increase in reassignments as the exam approaches. Overall, understanding patterns of reassignment is the focus of this paper, along with insights into how teachers were using the software component of CTAI.

@@

bbb <- addOverallState(data)%>%filter(!is.na(Curriculum) & State!='NJ')%>%
    group_by(state,Curriculum,overall,Yr)%>%
    summarize(n=n())
staten <- addOverallState(data)%>%filter(!is.na(Curriculum) & State!='NJ')%>%group_by(state,Yr)%>%summarize(n=n())

for(i in 1:nrow(bbb)) bbb$n[i] <- bbb$n[i]/staten$n[staten$state==bbb$state[i] & staten$Yr==bbb$Yr[i]]

ggplot(bbb,aes(Yr,n,fill=Curriculum,alpha=overall))+geom_col()+facet_grid(~state)+scale_alpha_manual(values=c(1,0.5))+scale_fill_manual(values=COLS)+scale_y_continuous(labels=percent)+labs(x='',y='% of Problems Worked',alpha='')

## ggplot(bbb,aes(Yr,n,color=bbb))+geom_col(show.legend=FALSE)+geom_col(aes(fill=Curriculum))+geom_col(aes(fill=overall),alpha=0.5)+facet_grid(~state)
##     scale_fill_manual(values=c('#8dd3c7','#8dd3c7','#ffffb3','#ffffb3','#bebada','white','black'))
##     geom_col(aes(Yr,n,fill=overall),alpha=0.5)+
##     #geom_col(aes(Yr,n,fill=overall))+#scale_fill_manual(values=c('white','black'))+
##     facet_grid(~state)

## omar <- par()$oma

## par(oma=c(0,2,0,0),mar=c( 3,.5,2,0))
## par(mfrow=c(1,9))
## for(st in levels(data$State)){
## #    cwy <- currWorkedSt%>%filter(State==st)%>%group_by(Curriculum,overall)%>%summarize(`Yr 1`=p[Year=='Year 1'],`Yr 2`=p[Year=='Year 2'])
##     cwy <- data%>%filter(!is.na(Curriculum) & State==st)%>%group_by(Curriculum,overall)%>%
##     summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
##     for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
##     cwy <- cwy[nrow(cwy):1,]
##     barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',50,NA),main=st,yaxt='n')
##     if(st==levels(data$State)[1]) axis(2,at=seq(0,1,0.1),labels=paste0(seq(0,100,10),'\\%'))
##     #box()
## }
## cwy <- data%>%filter(!is.na(Curriculum))%>%group_by(Curriculum,overall)%>%
##     summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
## for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
## cwy <- cwy[nrow(cwy):1,]
## barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',50,NA),main='Overall',yaxt='n')

## plot(1:10,type='n',xaxt='n',yaxt='n',bty='n')
## legend(-1.5,7.5, c("Bridge-to-\n Algebra",'Algebra I','$>$Alg. I','','Standard','Cust.'),
##        fill=c(COLS,NA,'#010101','#010101'),border=rep('white',6),
##        angle=c(NA,NA,NA,NA,NA,45),density=c(NA,NA,NA,NA,NA,50),xpd = TRUE, bty = "n",inset=c(0,0))

## par(oma=omar)
@
\caption{Percentage of worked problems coming from various levels
  (denoted by color, with Algebra II and Geometry bundled as ``$>$Algebra I''), from
  standard and customized variants, denoted by shading.}
\label{fig:curricula}
\end{figure}

@@

\begin{figure}
\centering
<<usageTime, fig.height=3,fig.width=6,dependson=c('data','setup','usageMedians')>>=

print(timeStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,time))+geom_boxplot(outlier.shape=NA)+
          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.time), width=0.2)+
          facet_grid(~state)+coord_cartesian(ylim=c(0,110))+labs(x='',y='Hours on CT Software per Student'))
                                        #ggsave('timeStatYear.jpg',width=6,height=3)

@@

<<supplementalFigs1,include=FALSE,dependson=c('data','setup','usageMedians')>>=
secStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,numSec))+geom_boxplot(outlier.shape=NA)+
          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.sec), width=0.2)+
          facet_grid(~state)+labs(x='',y='# of Sections Worked per Student')+coord_cartesian(ylim=c(0,200))
ggsave('secStateYear.jpg',width=6,height=3)

probStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,nprob))+geom_boxplot(outlier.shape=NA)+
          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.prob), width=0.2)+
          facet_grid(~state)+labs(x='',y='# of Problems Worked per Student')+coord_cartesian(ylim=c(0,2000))
ggsave('probStateYear.jpg',width=6,height=3)

@@

\begin{figure}
\centering
<<unitsWorked,dependson=c('usageMedians','data','setup'),fig.width=6,fig.height=3>>=
secByStud2 <-
  secByStud2 %>%
  group_by(state,Yr) %>%
  mutate(outlier = numUnit > quantile(numUnit,0.75,na.rm=TRUE) + IQR(numUnit,na.rm=TRUE) * 1.5) %>%
  ungroup

print(unitsStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,numUnit))+geom_boxplot(outlier.shape=NA)+geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier), width=0.2)+facet_grid(~state)+labs(x='',y='# Units Worked Per Student')+coord_cartesian(ylim=c(0,55)))
@
\caption{Boxplots of the number of units of Cognitive Tutor
  software each student worked, by year and state. Students
  working more than 55 units (\Sexpr{sum(secByStud[['numUnit']]>55,na.rm=TRUE)}) of \Sexpr{nrow(secByStud)}) and students
  with no usage data (\Sexpr{sum(is.na(secByStud[['numUnit']]))})
  were excluded.}
\label{fig:unitsByStud}
\end{figure}

@@

\section{Working Units in Order---Or Not}

@@

\begin{figure}
  \centering
<<unitsWorkedCust,dependson=c('data','setup','whichUnits'),fig.height=4,fig.width=6>>=
### by customized curriculum (at school level)
cust <- data%>%filter(Year=='Year 2')%>%group_by(schoolid2,state)%>%summarize(cust=mean(overall=='Customized',na.rm=T))%>%arrange(cust)
data$cust <- ifelse(data$schoolid2%in%cust$schoolid2[cust$cust>0.8],'Customized','Standard')

nstudCust <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(cust)%>%summarize(nstud=n_distinct(field_id))
unitLevelCust <- data%>%filter(Unit%in%units & Year=='Year 2' )%>%group_by(Unit,cust)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))

unitLevelCust$perWorked <- unitLevelCust$numWorked/nstudCust$nstud[match(unitLevelCust$cust,nstudCust$cust)]

unitLevelCust$Unit <- factor(unitLevelCust$Unit,levels=units)
levels(unitLevelCust$Unit) <- UnitName
ggplot(unitLevelCust,aes(x=Unit,y=perWorked,color=cust,group=cust))+geom_point()+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10),legend.position=c(.96,.82))+labs(x='',y='% Worked',color='Curriculum',title='Year 2')+scale_y_continuous(labels=percent)
@
\caption{The percentages of year-2 students with usage data who worked at
  least one problem from each unit in the Algebra I curriculum. The
  units are arranged in order for the standard curriculum. Students
  are divided between those attending schools using primarily a customized
  curriculum and those using primariy the standard Algebra I curriculum.}
\label{fig:unitsWorkedCust}
\end{figure}

Most of the variation in unit order was driven by the rise, in year 2,
of customized curricula.
Figure \ref{fig:unitsWorkedCust} divides year-2 students into those
attending schools using primarily a customized curriculum, and those
attending schools using primarily a standardized
curriculum.\footnote{At least
  \Sexpr{trunc(100*min(cust[['cust']][cust[['cust']]>0.8]))}\% of
  problems at ``Customized'' schools were from a customized curriculum,
  and at most
  \Sexpr{ceiling(100*max(cust[['cust']][cust[['cust']]<0.8]))}\% of
  problems at ``Standard'' schools were from a customized curriculum.}
Students using a standardized curriculum followed the standard
sequence---more or less---while students using customized curricula
did not.
That said, there were some order violations in the standard group:
specifically, more students worked problems from units ``4-Quadrant
Linear Graphs'' and ``Exponents'' than worked the preceding sections;
this suggests that some teachers used the reassignment tool to
prioritize particular topics.
Of course, this may have occurred in schools with customized
curriculua as well---a possibility we will discuss in the next
section.

\begin{figure}
  \centering
<<unitsBySchool,dependson=c('data','setup','whichUnits'),fig.height=7,fig.width=6>>=
### by school
nstudSch <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(schoolid2,state)%>%summarize(nstud=n_distinct(field_id))
unitLevelSch <- with(filter(data,Unit%in%units & Year=='Year 2'),expand.grid(Unit=unique(Unit),schoolid2=unique(schoolid2)))
unitLevelSch$state <- data$state[match(unitLevelSch$schoolid2,data$schoolid2)]
unitLevelSch$cust <- data$cust[match(unitLevelSch$schoolid2,data$schoolid2)]
unitLevelSch$numWorked <- with(filter(data,Year=='Year 2'),vapply(1:nrow(unitLevelSch),function(i)
    n_distinct(field_id[schoolid2==unitLevelSch$schoolid2[i] & Unit==unitLevelSch$Unit[i]]),1))

#unitLevelSch <- data%>%filter(Unit%in%units & Year=='Year 2')%>%group_by(Unit,schoolid2,cust,state)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))

unitLevelSch$perWorked <- unitLevelSch$numWorked/nstudSch$nstud[match(unitLevelSch$schoolid2,nstudSch$schoolid2)]

unitLevelSch$Unit <- factor(unitLevelSch$Unit,levels=units)
levels(unitLevelSch$Unit) <- UnitName

ggplot(filter(unitLevelSch,state!='NJ'),aes(x=Unit,y=perWorked,color=schoolid2,group=schoolid2,linetype=cust))+geom_point(size=.5)+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10))+
                                        #,legend.position=c(.96,.82))+
    facet_grid(state~.)+scale_y_continuous(labels=percent)+scale_color_discrete(guide=FALSE)+
    labs(x='',y='% Worked',linetype='Curriculum',title='Year 2')
@
\caption{The percentages of year-2 students with usage data in each
  school who worked at least one problem from each unit in the Algebra I curriculum. The
  units are arranged in order for the standard curriculum. Schools are
  classified as either using primarily customized curricula (solid
  line) or using primarily the standard Algebra I curriculum (dotted).}
\label{fig:unitsBySchool}
\end{figure}

Figure \ref{fig:unitsBySchool} further decomposes the year-2 results
by school and state, showing a large amount of variation between
states, as well as variation between schools within states.
In Texas, every school used customized curricula, most of which seem to
prioritize some of the same units, for instance, ``Linear Patterns,''
``Ind. Variables in Lin. Mod.,'' and
``Lin. Eq. w/ Var's Both Side.''
These topics might be emphasized on Texas standardized tests.
On the other hand, there was also variance between schools.
For instance, one school prioritized units ``2 Step Lin. Eq.''
and ``4−Quadrant Linear Graphs'' while nearly eliminating ``Linear Patterns.''

Between-school variation is evident in the other states, as well.
In four of the five Kentucky schools, nearly every student worked on
the first nine units; in the one Kentucky school that used a
customized curriculum, nearly every student worked on the first 13
units, omitted the 15th (``Lin. Mod. in General Form''), and worked on
the 16th and 17th (``Literal Eq.'' and ``Lin. Eq. w/ Var's Both Sides'').
In the remaining school, nearly every student worked on the first
section, but usage decreased rapidly from there.
Curiously, in one Michigan school which used the standard curriculum,
no students seem to have worked on the ``Lin. Mod. \& Ratios'' section.

If unit order and topic scaffolding are important to CT's mastery
learning mechanism, the wide variation in students' realized curricula
would seem to pose a problem.
The fact that the prescribed order was followed less in the second
year of the study, when CTAI was effective, than in the first year,
when it wasn't, suggests that the standard curriculum may play a
smaller role than one might otherwise imagine.

\section{Mastering the Material---Or Not}
The central idea behind mastery learning is that students progress
through the curriculum as they master skills.
In the context of CT, skills are clustered within sections, which are
in turn clustered within units.
Students progress from the current section to the next section after
mastering all of the current section's skills.
Ideally, students would master all of the skills in all of the
sections they work.

By default, the software operates by automatically moving students
from section to section based on the sequence of topics defined by the
curriculum they were currently enrolled in. In this
software-controlled sequencing, students ideally spend the time
necessary to learn the material of a section, are judged by the
software to have mastered the material, and then ``graduate'' to the
next section. Students who exhaust a section's material without
mastering its skills are “promoted” to the next section.
Students left uninterrupted follow the sequence of topics defined by
their curriculum, ideally graduating from topics by showing mastery.
However, teachers could modify a student’s path within the curriculum.
They could ``reassign'' students from their current sections to other
sections earlier or later in the intended sequence, including sections
they had worked on previously.
Finally, if the semester ends, or a student stops using CT for some
other reason, while in the middle of working through a
section, he or she will leave their ``Final'' section without
mastering all of its skills.
All in all, each CT section begun ends in one of four possible ways:
mastery, promotion, reassignment, or as the student's final section.

\begin{figure}
  \centering
<<overallStatus,dependson=c('data','setup'),fig.height=3,fig.width=6>>=
statusPerSec <- data%>%
    filter(!is.na(Curriculum) &!(state=='MI'&Curriculum=='Customized'&year==1))%>%
    group_by(field_id,Yr,state,section,Curriculum,status)%>%
        summarize(cp=any(status=='changed placement'))%>%group_by(state,Curriculum,Yr,field_id)%>%summarise(pcp=mean(cp,na.rm=TRUE),ncp=sum(cp,na.rm=TRUE))

statusOverall <-  data%>%filter(!is.na(status))%>%
    group_by(field_id,Yr,state,unit,section,Curriculum,overall)%>%summarize(status=max(status))

statusStateYr <- addOverallState(statusOverall) %>% group_by(state,Yr)%>%
    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))

levels(statusStateYr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Mastered='pgrad')

 ggplot(filter(statusStateYr,state!='NJ'),aes(Yr,value,fill=variable))+geom_col()+facet_grid(~state)+labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent)

@
\caption{The distributions of outcomes of worked sections, by state and across the
  entire sample, in the two study years.}
\label{fig:overallStatus}
\end{figure}

\begin{table}
 \begin{tabular}{rllllll|llllll}

&   \multicolumn{6}{c}{Year 1}&\multicolumn{6}{c}{Year 2}\\
<<statusTab,dependson='overallStatus',results='asis'>>=

tab1 <- dcast(subset(statusStateYr,Yr=='Yr 1'&state!='NJ'),variable~state)
tab2 <- dcast(subset(statusStateYr,Yr=='Yr 2'&state!='NJ'),variable~state)
tab <- cbind(tab1,tab2[-1])

tab$variable <- as.character(tab$variable)
cat('&')
cat(names(tab)[-1],sep='&')
cat('\\\\')
for(i in 1:nrow(tab)){
    cat(tab[i,1],round(unlist(tab[i,-1])*100),sep='&')
    cat('\\\\')
}

rownames(tab1) <- rownames(tab2) <- tab1$variable
tab1 <- as.matrix(tab1[,-1])
tab2 <- as.matrix(tab2[,-1])

\end{tabular}
\caption{Percentages of worked sections that ended in each of the
  four possible outcomes, across states and study years.}
\label{tab:overallStatus}
\end{table}

Figure \ref{fig:overallStatus} and Table \ref{tab:overallStatus} show
the proportions of worked sections in each state and study year that
ended with mastery, promotion, or reassignment, or which were the
student's final section.
In the first year, about
\Sexpr{round(mean(tab1['Mastered',c('TX','KY','MI','LA')])*100)}\% of
worked sections are mastered in every
state other than Connecticut.
Other than in Texas, about
\Sexpr{paste(round(range(tab1['Promoted',c('CT','KY','MI','LA')])*100),collapse='--')}\%
of sections end in promotion.
About \Sexpr{round(tab1['Reassigned','TX']*100)}
and \Sexpr{round(tab1['Reassigned','CT']*100)}\% of sections in Texas
and Connecticut end in reassignment, which is even rarer in the other states.

With the exception of Texas, sections in the second year tend to be
completed as they were in year 1.
In Texas, however, the percentage of sections ending in reassignment increased by a factor
of about about
\Sexpr{round(tab2['Reassigned','TX']/tab1['Reassigned','TX'])},
to about \Sexpr{round(tab2['Reassigned','TX']*100)}\%.
The proportion of Texas sections labeled ``Final'' increased as well,
most likely due to the decrease in usage.

Across states, just under
\Sexpr{round(tab1['Reassigned','Overall']*100)}\% of sections in year
1 eneded in reassignment, as did roughly
\Sexpr{round(tab2['Reassigned','Overall']*100)}\% in year 2.

\subsection{Section Mastery and Curriculum}
\begin{figure}
<<statusCur,dependson='overallStatus',fig.height=3,fig.width=6>>=
statusCurr <- statusOverall%>%filter(!is.na(Curriculum))%>% group_by(Curriculum,overall)%>%
    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
levels(statusCurr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Graduated='pgrad')
statusCurr$Curriculum <- factor(statusCurr$Curriculum,levels=rev(levels(statusCurr$Curriculum)))
print(ggplot(statusCurr,aes(overall,value,fill=variable))+geom_col()+facet_grid(~Curriculum)+labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent))

@
\caption{The distributions of outcomes of worked sections, by curriculum, in the two study years.}
\label{fig:statusCur}
\end{figure}

\section{Digging Deeper into Section Reassignment}
<<cpDat,dependson=c('data','setup'),include=FALSE>>=
secLev <- data%>%filter(is.finite(status) & is.finite(timestamp) &is.finite(date))%>%
    group_by(field_id,unit,section,Year,Yr,classid2,schoolid2)%>%
    summarize(startDate=min(date),endDate=max(date),startTime=min(timestamp),endTime=max(timestamp),state=state[1],
              status=max(status),Curriculum=Curriculum[1],overall=overall[1],version=version[1])%>%
    arrange(endDate)

### cp over time
secLev <- within(secLev,endMonth <- factor(month(secLev$endDate,TRUE,TRUE),levels=c('Aug','Sep','Oct','Nov','Dec','Jan','Feb','Mar','Apr','May','Jun','Jul')))

@
\subsection{When are Students Reassigned?}

\begin{figure}
  \centering
<<byMonth,dependson='cpDat',fig.height=3,fig.width=6>>=
secLevMonth <- secLev%>%group_by(endMonth,Year)%>%summarize(ncp=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
secLev$cp <- as.numeric(secLev$status=='changed placement')

ggplot(filter(secLevMonth,ncp>100),aes(endMonth,CPper,group=Year,color=Year,size=ncp))+geom_point()+geom_line(size=1)+
    ## geom_smooth(aes(as.numeric(endMonth)+day(endDate)/31-0.5,cp,group=Year,
    ##                 color=Year,size=1),
    ##             method = "glm", formula = y ~ splines::bs(x, 4),data=secLev,method.args=list(family='binomial'),
                                        #            show.legend=FALSE)+
scale_y_continuous(breaks=seq(0,0.05,.01),labels=percent)+
    coord_cartesian(ylim=c(0,0.05))+ labs(x='Month',y='% of Sections Ending in Reassignment')

@
\end{figure}

\begin{figure}
  \centering
<<byMonthState,dependson='cpDat',fig.height=3,fig.width=6>>=
secLevMonthSt <- secLev%>%filter(state%in%c('TX','KY','MI') & endMonth!='Jul')%>%group_by(endMonth,state,Year)%>%summarize(nsec=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
ggplot(secLevMonthSt,aes(endMonth,CPper,group=state,color=state,size=nsec))+geom_point()+facet_grid(Year~.)+
    geom_line(size=1)+
    ## geom_smooth(aes(as.numeric(endMonth)+day(endDate)/31-1.5,cp,group=state,
    ##                 color=state,size=1),
    ##             method = "glm",
    ##             formula = y ~ splines::ns(x, knots=seq(1,10,2)),
    ##             data=filter(secLev,state%in%c('TX','CT','NJ')& endMonth!='Jul'),
    ##             method.args=list(family='binomial'),
    ##             show.legend=FALSE)+
    scale_y_continuous(labels=percent)+
    labs(x='Month',y='% Sections Ending in Reassignment',size='Worked \n Sections',color='')

@
\end{figure}

\subsection{Does Reassignment Depend on Classmates?}
\begin{figure}
<<classmates,dependson='cpDat',fig.height=6,fig.width=7>>=
secLev$unitSectionClass <- paste(secLev$unit,secLev$section,secLev$classid2)
cpDat <- subset(secLev,status=='changed placement' & Curriculum=='Algebra I')
cpClasses <- unique(cpDat$unitSectionClass)
secLevCP <- filter(secLev,unitSectionClass%in%cpClasses)
levels(secLevCP$status) <- list(promGrad=c('promoted','graduated','final_or_incomplete'),cp='changed placement')
secLevCPsplit <- split(secLevCP[,c('field_id','endDate','status')],secLevCP$unitSectionClass)

ncp <- nrow(cpDat)
seqDat <- matrix(NA,nrow=ncp,ncol=6)
for(i in 1:ncp){
    if(i%%100==0) cat(round(i/ncp*100,2),'% ',sep='')
    cls <- secLevCPsplit[[cpDat$unitSectionClass[i]]]
    if(nrow(cls)==1){
        seqDat[i,] <- 0
        next
    }
    cls <- filter(cls,field_id!=cpDat$field_id[i])
    date <- cpDat$endDate[i]
    ord <- factor(ifelse(cls$endDate<date,'bf',
                  ifelse(cls$endDate==date,'same','after')),levels=c('bf','same','after'))
    clsSplit <- split(cls,list(ord,cls$status))
    seqDat[i,] <- vapply(clsSplit,nrow,1)
}
cat('\n')
colnames(seqDat) <- names(clsSplit)

classSize <- data%>%filter(!is.na(section))%>%group_by(classid2)%>%summarize(nstud=n_distinct(field_id))
cpDat$totClassmates <- classSize$nstud[match(cpDat$classid2,classSize$classid2)]

for(nn in names(clsSplit)) cpDat[[nn]] <- seqDat[,nn]/cpDat$totClassmates
cpDat$nw <- (cpDat$totClassmates-rowSums(seqDat))/cpDat$totClassmates
cpDat$total <- rowSums(seqDat)
cpDat$bf.promGrad <- cpDat$bf.promGrad+cpDat$same.promGrad

### plot them all

mmm <- with(subset(cpDat,totClassmates>15),max(table(state,Year)))

cpDat2 <- cpDat%>%filter(totClassmates>15)%>%arrange(bf.promGrad,bf.cp,same.cp,after.cp,after.promGrad)%>%group_by(Year,state)%>%mutate(cpid=seq(floor((mmm-n())/2)+1,floor((mmm-n())/2)+n()))%>%ungroup()

aaa <- melt(cpDat2,measure.vars=c('bf.promGrad','bf.cp','same.cp','after.cp','after.promGrad','nw'))
levels(aaa$variable) <- list(
    `Never Worked\n Section`='nw',
    `Mastered/Promoted\n Later`='after.promGrad',
    `Reassigned\n Later`='after.cp',
    `Reassigned\n Same Day`='same.cp',
    `Reassigned\n Before`='bf.cp',
    `Mastered/Promoted\n Before`='bf.promGrad')

#aaa$variable <- factor(aaa$variable,levels=rev(c('pGrad','pProm','pCPbf','pCPsame','pLater','pNW')))
ggplot(filter(aaa,state%in%c('TX','KY','MI')),aes(as.factor(cpid),value,fill=variable))+geom_col()+facet_grid(Year~state)+theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + scale_y_continuous(label=percent)+labs(fill='% of Classmates')+
    scale_fill_manual(values=c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e','#e6ab02'))

@
\end{figure}

\subsection{Where To?}
\begin{figure}
<<transition,fig.height=5,fig.width=6,dependson=c('data','setup')>>=
#### what's the next section each student works?
secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]))

secOrderCP <- subset(secOrder,status=='changed placement')

secOrderCP$Unit <- secOrderCP$unit
secOrderCP$Unit[grep('unit-conversions',secOrderCP$Unit)] <- 'unit-conversions'

secOrderCP$unitName <- factor(secOrderCP$Unit,levels=units)
levels(secOrderCP$unitName) <- UnitName

secOrderCP$nextUnit[grep('unit-conversions',secOrderCP$nextUnit)] <- 'unit-conversions'

secOrderCP$nextUnitName <- factor(secOrderCP$nextUnit,levels=units)
levels(secOrderCP$nextUnitName) <- UnitName

trans <- with(secOrderCP,table(unitName,nextUnitName))

trans <- trans[rowSums(trans)>50,rowSums(trans)>50]
trans <- trans[rowSums(trans)>50,rowSums(trans)>50]

transitionPlot(trans)

nsec <- nrow(trans)
